# Afib - cancer proteomic analysis using prevalence and 41270 field
# AFib as control, cancer as cases

---
title: "Cancer-prevalent AFib as Controls Proteomics"
output: html_document
date: "`r Sys.Date()`"
---
### Load in libraries
```{r}
library(readr)
library(dplyr)
library(data.table)
library(ukbtools)
library(purrr)
library(janitor)
library(tidyr)
library(ggplot2)
library(survival)
library(broom)
library(purrr)
library(ggrepel)
library(ggbreak)
library(gridExtra)
library(grid)
library(R.utils)
library(TwoSampleMR)
library(genetics.binaRies)
library(ieugwasr)
library(stringr)
library(jsonlite)
library(forcats)

```

### Load in proteins file, covariates and data dict
```{r}
olink_df <- read_tsv("/mnt/project/Proteomics/olink_npx_clean_imputed.tsv")
covariates_named <- read_tsv("/mnt/project/Proteomics/covariates_named.tsv")
dict <- read_csv("/mnt/project/Proteomics/raw/ddd/proteomics_baseline.data_dictionary.csv")

```
### Let's combine the proteomics and covariates datasets

```{r}
merged_df <- inner_join(olink_df, covariates_named, by = "eid")

```

### Exclude relatives
```{r}

rel_path <- "/mnt/project/Bulk/Genotype Results/Genotype calls/ukb_rel.dat"  # update to actual path if needed
kin_df <- data.table::fread(rel_path)

#Get vector of our IDs
eid_vec <- merged_df$eid

set.seed(1)
to_remove <- ukb_gen_samples_to_remove(kin_df, ukb_with_data = eid_vec, cutoff = 0.0884)


#Filter
merged_df <- merged_df %>%
  filter(!eid %in% to_remove)

dim(merged_df)
```

### Get the relevant fields using data field 41270 for ICD
```{r}
field_list <- c("participant.eid", "participant.p41270")
writeLines(field_list, "~/afib/data/raw/icd10_field_list_41270.txt")
```

### Extract the fields from RAP
```{r}
field_list_file <- "~/afib/data/raw/icd10_field_list_41270.txt"
output_file <- "~/afib/data/raw/icd10_diag_41270.tsv"

cmd <- glue::glue(
  "dx extract_dataset /Proteomics/proteomics_baseline ",
  "--fields \"$(paste -sd, {field_list_file})\" ",
  "--delimiter '\\t' ",
  "--output {output_file}"
)

system(cmd)

```


### Load metadata

```{r}
# Load extracted ICD-10 data
icd_df <- read_tsv("~/afib/data/raw/icd10_diag_41270.tsv")

# Parse and unnest ICD codes
icd_long <- icd_df %>%
  filter(!is.na(participant.p41270), participant.p41270 != "") %>%
  mutate(icd_list = map(participant.p41270, ~ fromJSON(.) %>% as.character())) %>%
  select(participant.eid, icd_list) %>%
  unnest(icd_list) %>%
  rename(icd10_code = icd_list)
```

### AF metadata

```{r}
afib_metadata <- icd_long %>%
  filter(str_detect(icd10_code, "^I48")) %>%
  distinct(eid = participant.eid) %>%
  mutate(afib_status = 1)
```


### Cancer metadata

```{r}
# Load map of defined cancer groupings
defined_cancer_map <- tribble(
  ~icd_prefix, ~defined_group,
  "C18", "colorectal",
  "C19", "colorectal",
  "C20", "colorectal",
  "C50", "breast",
  "C61", "prostate",
  "C34", "lung",
  "C22", "liver",
  "C64", "kidney",
  "C73", "thyroid",
  "C67", "bladder",
  "C56", "ovary",
  "C54", "uterus",
  "C71", "brain",
  "C25", "pancreas",
  "C15", "esophagus",
  "C81", "hodgkin",
  "C82", "nhl",
  "C83", "nhl",
  "C85", "nhl",
  "C90", "myeloma",
  "C91", "leukemia",
  "C92", "leukemia",
  "C94", "leukemia"
)

# Step 2a: Filter cancer ICDs and annotate
cancer_icd_annotated <- icd_long %>%
  filter(str_detect(icd10_code, "^C[0-9]{2}")) %>%
  distinct(participant.eid, icd10_code) %>%
  mutate(icd_prefix = substr(icd10_code, 1, 3)) %>%
  left_join(defined_cancer_map, by = "icd_prefix") %>%
  mutate(defined_group = coalesce(defined_group, "other"))

# Step 2b: Create one-row-per-person metadata
cancer_metadata <- cancer_icd_annotated %>%
  group_by(participant.eid) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(cancer_status = 1) %>%
  select(eid = participant.eid, cancer_status, icd_group = defined_group, cancer_icd = icd10_code)
```

### Combine

```{r}
working_df <- merged_df %>%
  left_join(afib_metadata, by = "eid") %>%
  left_join(cancer_metadata, by = "eid") %>%
  mutate(
    afib_status = replace_na(afib_status, 0),
    cancer_status = replace_na(cancer_status, 0),
    icd_group = replace_na(icd_group, "no_cancer"),
    cancer_icd = replace_na(cancer_icd, "None")
  )

working_df %>%
  count(afib_status, cancer_status) %>%
  mutate(label = case_when(
    afib_status == 1 & cancer_status == 1 ~ "AFib + Cancer",
    afib_status == 1 & cancer_status == 0 ~ "AFib only",
    afib_status == 0 & cancer_status == 1 ~ "Cancer only",
    TRUE ~ "Neither"
  )) %>%
  select(label, n) %>%
  arrange(desc(n)) %>%
  knitr::kable(caption = "Final Cohort Breakdown")

```
### Summarize
```{r}
icd_summary <- working_df %>%
  filter(cancer_status == 1) %>%  # only participants with cancer
  group_by(icd_group) %>%
  summarise(
    Total = n(),
    AFib = sum(afib_status == 1),
    AFib_Rate = round(AFib / Total, 3),
    .groups = "drop"
  ) %>%
  arrange(desc(Total))


knitr::kable(
  icd_summary,
  caption = "AFib Counts and Rates Within Each Cancer Group",
  col.names = c("Cancer Group", "Total with Cancer", "AFib Cases", "AFib Rate")
)

cancer_status_summary <- working_df %>%
  count(cancer_status) %>%
  mutate(
    cancer_status = if_else(cancer_status == 1, "Cancer", "No Cancer")
  )

knitr::kable(
  cancer_status_summary,
  caption = "Total Participants by Cancer Status",
  col.names = c("Cancer Status", "n")
)

```

### Impute missing covariates

```{r}
# covariates to work off of
covariates <- c("bmi", "townsend_index", "alcohol_status", 
                "sbp_1", "dbp_1", 
                "total_cholesterol", "ldl_cholesterol", "hdl_cholesterol", 
                "triglycerides", "creatinine")

working_df <- working_df %>%
  mutate(across(all_of(covariates),
                .fns = ~ as.integer(is.na(.)),
                .names = "{.col}_missing"))

###  Step 2: Fit linear models to predict missing values ####

predictors <- c("sex", "age", "ethnicity", paste0("PC", 1:10))

lm_models <- map(covariates, function(var) {
  formula <- as.formula(paste(var, "~", paste(predictors, collapse = " + ")))
  lm(formula, data = working_df)
})
names(lm_models) <- covariates


###  Step 3: Generate predictions ####
predictions <- map2_dfc(
  covariates, lm_models,
  ~ tibble(!!paste0(.x, "_predicted") := predict(.y, newdata = working_df))
)

working_df <- bind_cols(working_df, predictions)


###  Step 4: Impute missing values with predicted ####
for (var in covariates) {
  working_df[[paste0(var, "_final")]] <- ifelse(
    is.na(working_df[[var]]),
    working_df[[paste0(var, "_predicted")]],
    working_df[[var]]
  )
}

###  Missingness summary before impu````````````````````````````tation ####

missing_summary <- working_df %>%
  select(all_of(covariates)) %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "PercentMissing") %>%
  arrange(desc(PercentMissing))

knitr::kable(missing_summary, digits = 2, caption = "Percent Missingness Before Imputation")

ggplot(missing_summary, aes(x = reorder(Variable, PercentMissing), y = PercentMissing)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Covariate Missingness", x = "", y = "% Missing")

write_csv(missing_summary, "~/afib/outputs/proteomics/covariate_missingness_summary.csv")

```

### Let's save the data now

```{r}
###  Save final cleaned + imputed dataset ####

output_path <- "~/afib/data/processed/ukb_proteomics_afib_cleaned.tsv"

# Write the final table as tab-delimited without row names or quotes
write.table(working_df, file = output_path, sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)

```

### Quick metadata

```{r}
###  Quick metadata for table 2 ####

table2_df <- working_df %>%
  mutate(
    study_group = case_when(
      afib_status == 1 & cancer_status == 0 ~ "AF-Only",
      afib_status == 1 & cancer_status == 1 ~ "AF+PanCancer",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(study_group))

# Add AF+Lung and AF+NHL as separate groups
# (these participants also appear in AF+PanCancer, which is fine — they're not mutually exclusive)
lung_df <- working_df %>%
  filter(afib_status == 1, icd_group == "lung") %>%
  mutate(study_group = "AF+Lung")

nhl_df <- working_df %>%
  filter(afib_status == 1, icd_group == "nhl") %>%
  mutate(study_group = "AF+NHL")

table2_df <- bind_rows(table2_df, lung_df, nhl_df)

# Now summarize
table2_summary <- table2_df %>%
  mutate(
    male = as.integer(sex == 1),
    white = as.integer(ethnicity %in% c(1, 1001, 1002, 1003))
  ) %>%
  group_by(study_group) %>%
  summarise(
    n = n(),
    age_mean = round(mean(age, na.rm = TRUE), 1),
    age_sd = round(sd(age, na.rm = TRUE), 1),
    male_n = sum(male, na.rm = TRUE),
    male_pct = round(100 * mean(male, na.rm = TRUE), 1),
    white_n = sum(white, na.rm = TRUE),
    white_pct = round(100 * mean(white, na.rm = TRUE), 1),
    bmi_mean = round(mean(bmi, na.rm = TRUE), 1),
    bmi_sd = round(sd(bmi, na.rm = TRUE), 1),
    .groups = "drop"
  )

print(table2_summary)

```

### Logistic regression for composite cancer

```{r}

# Define covariates and collapse into string
covariates <- c("age", "sex", paste0("PC", 1:10), "ethnicity")
cov_str <- paste(covariates, collapse = " + ")

# Define proteins
protein_vars <- setdiff(names(olink_df), "eid")

# Subset to AFib participants only
afib_df <- working_df %>%
  filter(afib_status == 1)

# Container
all_logit_results <- list()

# Loop through all proteins
logit_results <- map_dfr(protein_vars, function(prot) {
  formula_str <- paste0("cancer_status ~ ", prot, " + ", cov_str)
  
  tryCatch({
    fit <- glm(as.formula(formula_str), data = afib_df, family = binomial)
    tidy(fit) %>%
      filter(term == prot) %>%
      mutate(protein = prot)
  }, error = function(e) NULL)
})

# Format
logit_results_clean <- logit_results %>%
  select(term, estimate, std.error, statistic, p.value, protein) %>%
  rename(
    Log_OR = estimate,
    SE = std.error,
    Z = statistic,
    P = p.value
  ) %>%
  mutate(
    OR = exp(Log_OR),
    CI_Lower = exp(Log_OR - 1.96 * SE),
    CI_Upper = exp(Log_OR + 1.96 * SE),
    Group = "PanCancer",
    P_adj = p.adjust(P, method = "fdr")
  )

# Save
write_csv(logit_results_clean, "~/afib/outputs/proteomics/afib_cancer/afib_control/logit_pancancer.csv")

# Visualization
sample_n <- working_df %>%
  filter(afib_status == 1) %>%
  count(cancer_status) %>%
  mutate(label = if_else(cancer_status == 1, "AFib + Cancer", "AFib Only")) %>%
  select(label, n)

# Convert to named values
n_cancer <- sample_n$n[sample_n$label == "AFib + Cancer"]
n_afib_only <- sample_n$n[sample_n$label == "AFib Only"]

# Prepare volcano dataframe
volcano_df <- logit_results_clean %>%
  mutate(
    neg_log_p = -log10(P_adj),
    direction = case_when(
      P_adj >= 0.05 ~ "Neutral",
      Log_OR > 0 ~ "Up (Risk ↑)",
      Log_OR < 0 ~ "Down (Protective)"
    ),
    label = ifelse(P_adj < 0.05 & abs(Log_OR) > 0.25, protein, NA)  # customize threshold here
  )

# Plot
ggplot(volcano_df, aes(x = Log_OR, y = neg_log_p, color = direction)) +
  geom_point(alpha = 0.9, size = 1.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "darkred") +
  geom_text_repel(aes(label = label), size = 4, max.overlaps = 25) +
  scale_color_manual(
    values = c(
      "Up (Risk ↑)" = "#D55E00",
      "Down (Protective)" = "#0072B2",
      "Neutral" = "grey70"
    )
  ) +
  labs(
  title = "Proteins Associated with Pan-Cancer in AFib-Only Patients",
  subtitle = paste0("AFib + Cancer (n = ", n_cancer, "), AFib Only (n = ", n_afib_only, ")"),
  x = "log(Odds Ratio)",
  y = expression(-log[10]~"(FDR-Adjusted P-Value)"),
  color = "Direction"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_cartesian(ylim = c(0, 20))  # adjust as needed


# Prepare volcano dataframe with adjusted pval
volcano_df <- logit_results_clean %>%
  mutate(
    neg_log_p = -log10(P_adj),
    direction = case_when(
      P_adj >= 0.1 ~ "Neutral",
      Log_OR > 0 ~ "Up (Risk ↑)",
      Log_OR < 0 ~ "Down (Protective)"
    ),
    label = ifelse(P_adj < 0.1, protein, NA)  # threshold adjusted here too
  )

# Plot
ggplot(volcano_df, aes(x = Log_OR, y = neg_log_p, color = direction)) +
  geom_point(alpha = 0.9, size = 1.5) +
  geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "darkred") +
  geom_text_repel(aes(label = label), size = 4, max.overlaps = 25) +
  scale_color_manual(
    values = c(
      "Up (Risk ↑)" = "#D55E00",
      "Down (Protective)" = "#0072B2",
      "Neutral" = "grey70"
    )
  ) +
  labs(
  title = "Proteins Associated with Pan-Cancer in AFib-Only Patients",
  subtitle = paste0("AFib + Cancer (n = ", n_cancer, "), AFib Only (n = ", n_afib_only, ")"),
  x = "log(Odds Ratio)",
  y = expression(-log[10]~"(FDR-Adjusted P-Value)"),
  color = "Direction"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_cartesian(ylim = c(0, 2))  # adjust as needed


```

### ok, loop thru diff cancers now

```{r}
#cancers of interest
cancers <- c("prostate", "breast", "colorectal", "lung", "bladder", "esophagus", "nhl")

covariates <- c("age", "sex", paste0("PC", 1:10), "ethnicity")
cov_str <- paste(covariates, collapse = " + ")
protein_vars <- setdiff(names(olink_df), "eid")


# Filter to only AFib patients
afib_df <- working_df %>%
  filter(afib_status == 1)

# Container
all_results <- list()

for (cancer in cancers) {
  message("Analyzing cancer group: ", cancer)
  
  # Create binary outcome: 1 = AFib + specific cancer, 0 = AFib-only (no cancer)
  group_df <- afib_df %>%
    mutate(case = if_else(icd_group == cancer, 1, 0)) %>%
    filter(case == 1 | cancer_status == 0)  # Keep only AFib-only or AFib + this cancer
  
  # Skip if not enough cancer cases
  n_cancer <- sum(group_df$case == 1)
  if (n_cancer < 10) {
    message("  Skipping ", cancer, ": only ", n_cancer, " cancer cases.")
    next
  }

  # Run regression for each protein
  group_results <- map_dfr(protein_vars, function(prot) {
    formula_str <- paste0("case ~ ", prot, " + ", cov_str)
    tryCatch({
      fit <- glm(as.formula(formula_str), data = group_df, family = binomial)
      tidy(fit) %>%
        filter(term == prot) %>%
        mutate(protein = prot)
    }, error = function(e) NULL)
  })

  # Format results
  group_results_clean <- group_results %>%
    select(term, estimate, std.error, statistic, p.value, protein) %>%
    rename(
      Log_OR = estimate,
      SE = std.error,
      Z = statistic,
      P = p.value
    ) %>%
    mutate(
      OR = exp(Log_OR),
      CI_Lower = exp(Log_OR - 1.96 * SE),
      CI_Upper = exp(Log_OR + 1.96 * SE),
      Group = cancer,
      P_adj = p.adjust(P, method = "fdr")
    )

  # Save
  all_results[[cancer]] <- group_results_clean
  write_csv(group_results_clean, paste0("~/afib/outputs/proteomics/afib_cancer/afib_control/logit_", cancer, ".csv"))
}

# Combine all
combined_afibonly_results <- bind_rows(all_results)
write_csv(combined_afibonly_results, "~/afib/outputs/proteomics/afib_cancer/afib_control/logit_combined.csv")
combined_afibonly_results <- read_csv("~/afib/outputs/proteomics/afib_cancer/afib_control/logit_combined.csv")

# Get unique cancer groups
groups <- unique(combined_afibonly_results$Group)

# Get sample sizes
sample_n_df <- working_df %>%
  filter(afib_status == 1) %>%
  mutate(group = if_else(icd_group %in% groups, icd_group, "afib_only")) %>%
  group_by(group) %>%
  summarise(n = n(), .groups = "drop")

pdf("~/afib/outputs/proteomics/afib_cancer/afib_control/volcano_plots_all.pdf",
    width = 7, height = 5)

# Loop through and plot
for (grp in groups) {
  message("Plotting: ", grp)
  
  n_total <- sample_n_df$n[sample_n_df$group == grp]
  n_afib_only <- sample_n_df$n[sample_n_df$group == "afib_only"]
  
  volcano_df <- combined_afibonly_results %>%
    filter(Group == grp) %>%
    mutate(
      neg_log_p = -log10(P_adj),
      direction = case_when(
        P_adj >= 0.05 ~ "Neutral",
        Log_OR > 0 ~ "Positive Assocation (OR > 1)",
        Log_OR < 0 ~ "Negative Assocation (OR < 1)"
      ),
      label = ifelse(P_adj < 0.05 & abs(Log_OR) > 0.25, protein, NA)
    )
  
  # Plot
  p <- ggplot(volcano_df, aes(x = Log_OR, y = neg_log_p, color = direction)) +
    geom_point(alpha = 0.9, size = 1.5) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "darkred") +
    geom_text_repel(aes(label = label), size = 4, max.overlaps = 25) +
    scale_color_manual(
      values = c(
         "Positive Assocation (OR > 1)" = "#D55E00",
         "Negative Assocation (OR < 1)" = "#0072B2",
        "Neutral" = "grey70"
      )
    ) +
    labs(
    title = paste("AFib +", grp, "vs AFib Only"),
    subtitle = paste0("n = ", n_total, " with ", grp, ", n = ", n_afib_only, " AFib-only"),
    x = "log(Odds Ratio)",
    y = expression(-log[10]~"(FDR-Adjusted P-Value)"),
    color = "Direction"
     ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5)
    ) +
    coord_cartesian(ylim = c(0, 20))  # Adjust as needed
  
  print(p)
}
  
dev.off()

```
### Ok, cancer control
```{r}
covariates <- c("age", "sex", paste0("PC", 1:10), "ethnicity")
cov_str <- paste(covariates, collapse = " + ")

# 7 selected groups (from your cleaned icd_group column)
cancer_groups <- c("prostate", "breast", "colorectal", "lung", "bladder", "esophagus", "nhl")

# Protein variables
protein_vars <- setdiff(names(olink_df), "eid")
cov_str <- paste(c("age", "sex", paste0("PC", 1:10), "ethnicity"), collapse = " + ")

# Store results
all_results <- list()

# First: Pan-cancer (all cancer patients)
message("Running pan-cancer AFib model")

cancer_df <- working_df %>% filter(cancer_status == 1)

pan_results <- map_dfr(protein_vars, function(prot) {
  formula_str <- paste0("afib_status ~ ", prot, " + ", cov_str)
  tryCatch({
    fit <- glm(as.formula(formula_str), data = cancer_df, family = binomial)
    tidy(fit) %>%
      filter(term == prot) %>%
      mutate(protein = prot)
  }, error = function(e) NULL)
})


pan_clean <- pan_results %>%
  select(term, estimate, std.error, statistic, p.value, protein) %>%
  rename(Log_OR = estimate, SE = std.error, Z = statistic, P = p.value) %>%
  mutate(
    OR = exp(Log_OR),
    CI_Lower = exp(Log_OR - 1.96 * SE),
    CI_Upper = exp(Log_OR + 1.96 * SE),
    Group = "pan_cancer",
    P_adj = p.adjust(P, method = "fdr")
  )

all_results[["pan_cancer"]] <- pan_clean
write_csv(pan_clean, "~/afib/outputs/proteomics/afib_cancer/cancer_control/logit_pancancer.csv")


# Now loop through each cancer group
for (group in cancer_groups) {
  message("Analyzing cancer group: ", group)
  
  group_df <- working_df %>%
    filter(icd_group == group)
  
  n_afib <- sum(group_df$afib_status == 1)
  if (n_afib < 10) {
    message("  Skipping due to insufficient AFib cases (n = ", n_afib, ")")
    next
  }

  group_results <- map_dfr(protein_vars, function(prot) {
    formula_str <- paste0("afib_status ~ ", prot, " + ", cov_str)
    tryCatch({
      fit <- glm(as.formula(formula_str), data = group_df, family = binomial)
      tidy(fit) %>%
        filter(term == prot) %>%
        mutate(protein = prot)
    }, error = function(e) NULL)
  })

  group_clean <- group_results %>%
    select(term, estimate, std.error, statistic, p.value, protein) %>%
    rename(Log_OR = estimate, SE = std.error, Z = statistic, P = p.value) %>%
    mutate(
      OR = exp(Log_OR),
      CI_Lower = exp(Log_OR - 1.96 * SE),
      CI_Upper = exp(Log_OR + 1.96 * SE),
      Group = group,
      P_adj = p.adjust(P, method = "fdr")
    )

  all_results[[group]] <- group_clean
  write_csv(group_clean, paste0("~/afib/outputs/proteomics/afib_cancer/cancer_control/logit_", group, ".csv"))
}

# Combine all results
combined_results <- bind_rows(all_results)
write_csv(combined_results, "~/afib/outputs/proteomics/afib_cancer/cancer_control/logit_combined.csv")

combined_cancercontrol_results <- read_csv("~/afib/outputs/proteomics/afib_cancer/cancer_control/logit_combined.csv")

# Sample size table for each cancer group
groups <- unique(combined_cancercontrol_results$Group)

sample_n_df <- working_df %>%
  filter(cancer_status == 1) %>%
  group_by(group = icd_group, afib_status) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = afib_status, values_from = n, values_fill = 0) %>%
  rename(n_no_afib = `0`, n_afib = `1`) %>%
  bind_rows(
    working_df %>%
      filter(cancer_status == 1) %>%
      summarise(n_no_afib = sum(afib_status == 0),
                n_afib = sum(afib_status == 1)) %>%
      mutate(group = "pan_cancer")
  )

plot_list <- list()

for (i in seq_along(groups)) {
  grp <- groups[i]
  
  n_afib <- sample_n_df$n_afib[sample_n_df$group == grp]
  n_cancer_only <- sample_n_df$n_no_afib[sample_n_df$group == grp]
  
  volcano_df <- combined_cancercontrol_results %>%
    filter(Group == grp) %>%
    mutate(
      neg_log_p = -log10(P_adj),
      direction = case_when(
        P_adj >= 0.05 ~ "Neutral",
        Log_OR > 0 ~ "Positive Association (OR > 1)",
        Log_OR < 0 ~ "Negative Association (OR < 1)"
      ),
      label = ifelse(P_adj < 0.05 & abs(Log_OR) > 0.25, protein, NA)
    )
  
  p <- ggplot(volcano_df, aes(x = Log_OR, y = neg_log_p, color = direction)) +
    geom_point(alpha = 0.9, size = 1.5) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "darkred") +
    geom_text_repel(aes(label = label), size = 4, max.overlaps = 25) +
    scale_color_manual(values = c(
      "Positive Association (OR > 1)" = "#D55E00",
      "Negative Association (OR < 1)" = "#0072B2",
      "Neutral" = "grey70"
    )) +
    labs(
      title = paste("AF +", grp, "vs", grp, "Only"),
      subtitle = paste0("n = ", n_afib, " with AF, n = ", n_cancer_only, " cancer-only"),
      x = "log(Odds Ratio)",
      y = expression(-log[10]~"(FDR-Adjusted P-Value)"),
      color = "Direction"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5)
    ) +
    coord_cartesian(ylim = c(0, 20))
  
  plot_list[[i]] <- p
}

pdf("~/afib/outputs/proteomics/afib_cancer/cancer_control/volcano_plots_all.pdf",
    width = 7, height = 5)
for (p in plot_list) {
  print(p)
}
dev.off()


```

### Add in core afib proteins as PCs
```{r}

# Load in significant proteins
cox_df <- read_csv("~/afib/outputs/proteomics/mono_afib/cox_results.csv")

core_afib_proteins <- cox_df %>%
  filter(P_adj < 0.05) %>%
  arrange(P_adj) %>%
  pull(Protein) %>%
  unique()

# Perform PCA on scaled matrix of core AFib proteins
core_mat <- working_df %>%
  select(all_of(core_afib_proteins)) %>%
  scale()

core_pcs <- prcomp(core_mat, center = TRUE, scale. = TRUE)

# Add first 5 PCs to main dataframe
df_pc <- working_df %>%
  bind_cols(as_tibble(core_pcs$x[, 1:5]) %>% set_names(paste0("corePC", 1:5)))

# Updated covariate list
covariates <- c(
  "age", "sex",
  paste0("PC", 1:10),
  "ethnicity",
  paste0("corePC", 1:5)
)
cov_str <- paste(covariates, collapse = " + ")

# Set protein variables
protein_vars <- setdiff(names(olink_df), "eid")

# Cancer groups to loop through
cancer_groups <- c("pan_cancer", "prostate", "breast", "colorectal", "lung", "bladder", "esophagus", "nhl")

# Container for results
all_logit_results <- list()

for (group in cancer_groups) {
  message("Analyzing cancer group: ", group)
  
  # Filter for each cancer group
  group_df <- if (group == "pan_cancer") {
    df_pc %>% filter(cancer_status == 1)
  } else {
    df_pc %>% filter(icd_group == group)
  }
  
  n_afib <- sum(group_df$afib_status == 1)
  n_control <- sum(group_df$afib_status == 0)
  
  if (n_afib < 10 | n_control < 10) {
    message("  Skipping due to small sample size (AFib = ", n_afib, ", Control = ", n_control, ")")
    next
  }
  
  # Loop through proteins
  group_results <- map_dfr(protein_vars, function(prot) {
    formula_str <- paste0("afib_status ~ ", prot, " + ", cov_str)
    
    tryCatch({
      fit <- glm(as.formula(formula_str), data = group_df, family = binomial)
      tidy(fit) %>%
        filter(term == prot) %>%
        mutate(protein = prot)
    }, error = function(e) NULL)
  })

  # Format
  if (nrow(group_results) > 0) {
    group_results_clean <- group_results %>%
      select(term, estimate, std.error, statistic, p.value, protein) %>%
      rename(
        Log_OR = estimate,
        SE = std.error,
        Z = statistic,
        P = p.value
      ) %>%
      mutate(
        OR = exp(Log_OR),
        CI_Lower = exp(Log_OR - 1.96 * SE),
        CI_Upper = exp(Log_OR + 1.96 * SE),
        Group = group,
        P_adj = p.adjust(P, method = "fdr")
      )
    
    all_logit_results[[group]] <- group_results_clean
    write_csv(group_results_clean, paste0("~/afib/outputs/proteomics/afib_cancer/cancer_control/logit_pcadj_", group, ".csv"))
  }
}

# Combine
combined_canceronly_pcadj_results <- bind_rows(all_logit_results)
write_csv(combined_canceronly_pcadj_results, "~/afib/outputs/proteomics/afib_cancer/cancer_control/logit_combined.csv")



# Sample sizes
sample_n_df <- df_pc %>%
  filter(cancer_status == 1) %>%
  group_by(group = icd_group, afib_status) %>%
  summarise(n = n(), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = afib_status, values_from = n, values_fill = 0) %>%
  rename(n_no_afib = `0`, n_afib = `1`) %>%
  mutate(n_total = n_no_afib + n_afib) %>%
  bind_rows(
    df_pc %>%
      filter(cancer_status == 1) %>%
      summarise(n_no_afib = sum(afib_status == 0),
                n_afib = sum(afib_status == 1)) %>%
      mutate(group = "pan_cancer", n_total = n_no_afib + n_afib)
  )

# Plot
groups <- unique(combined_canceronly_pcadj_results$Group)

for (grp in groups) {
  message("Plotting: ", grp)
  
  n_afib <- sample_n_df$n_afib[sample_n_df$group == grp]
  n_cancer_only <- sample_n_df$n_no_afib[sample_n_df$group == grp]
  
  volcano_df <- combined_canceronly_pcadj_results %>%
    filter(Group == grp) %>%
    mutate(
      neg_log_p = -log10(P_adj),
      direction = case_when(
        P_adj >= 0.05 ~ "Neutral",
        Log_OR > 0 ~ "Up (Risk ↑)",
        Log_OR < 0 ~ "Down (Protective)"
      ),
      label = ifelse(P_adj < 0.05 & abs(Log_OR) > 0.25, protein, NA)
    )
  
  # Plot
  p <- ggplot(volcano_df, aes(x = Log_OR, y = neg_log_p, color = direction)) +
    geom_point(alpha = 0.9, size = 1.5) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "darkred") +
    geom_text_repel(aes(label = label), size = 4, max.overlaps = 25) +
    scale_color_manual(
      values = c(
        "Up (Risk ↑)" = "#D55E00",
        "Down (Protective)" = "#0072B2",
        "Neutral" = "grey70"
      )
    ) +
    labs(
      title = paste("AFib +", grp, "vs", grp, "Only"),
      subtitle = paste0("n = ", n_afib, " with AFib, n = ", n_cancer_only, " cancer-only"),
      x = "log(Odds Ratio)",
      y = expression(-log[10]~"(FDR-Adjusted P-Value)"),
      color = "Direction"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5)
    ) +
    coord_cartesian(ylim = c(0, 20))

  print(p)
}


```



### mono_cancer
```{r}
covariates <- c("age", "sex", paste0("PC", 1:10), "ethnicity")
cov_str <- paste(covariates, collapse = " + ")

# 7 selected groups (from your cleaned icd_group column)
cancer_groups <- c("prostate", "breast", "colorectal", "lung", "bladder", "esophagus", "nhl")

# Protein variables
protein_vars <- setdiff(names(olink_df), "eid")
cov_str <- paste(c("age", "sex", paste0("PC", 1:10), "ethnicity"), collapse = " + ")

# Store results
all_results <- list()

# First: Pan-cancer (all cancer patients)
message("Running pan-cancer AFib model")


pan_results <- map_dfr(protein_vars, function(prot) {
  formula_str <- paste0("cancer_status ~ ", prot, " + ", cov_str)
  tryCatch({
    fit <- glm(as.formula(formula_str), data = working_df, family = binomial)
    tidy(fit) %>%
      filter(term == prot) %>%
      mutate(protein = prot)
  }, error = function(e) NULL)
})


pan_clean <- pan_results %>%
  select(term, estimate, std.error, statistic, p.value, protein) %>%
  rename(Log_OR = estimate, SE = std.error, Z = statistic, P = p.value) %>%
  mutate(
    OR = exp(Log_OR),
    CI_Lower = exp(Log_OR - 1.96 * SE),
    CI_Upper = exp(Log_OR + 1.96 * SE),
    Group = "pan_cancer",
    P_adj = p.adjust(P, method = "fdr")
  )

all_results[["pan_cancer"]] <- pan_clean
write_csv(pan_clean, "~/afib/outputs/proteomics/mono_cancer/logit_pancancer.csv")


# Now loop through each cancer group
for (group in cancer_groups) {
  message("Analyzing cancer group: ", group)

  group_df <- working_df %>%
    filter(icd_group == group | cancer_status == 0) %>%
    mutate(
      cancer_label = if_else(icd_group == group, 1, 0)
    )

  n_case <- sum(group_df$cancer_label == 1)
  n_control <- sum(group_df$cancer_label == 0)

  if (n_case < 10 | n_control < 10) {
    message("  Skipping due to small group sizes: n_case = ", n_case, ", n_control = ", n_control)
    next
  }

  group_results <- map_dfr(protein_vars, function(prot) {
    formula_str <- paste0("cancer_label ~ ", prot, " + ", cov_str)
    tryCatch({
      fit <- glm(as.formula(formula_str), data = group_df, family = binomial)
      tidy(fit) %>%
        filter(term == prot) %>%
        mutate(protein = prot)
    }, error = function(e) NULL)
  })

  group_clean <- group_results %>%
    select(term, estimate, std.error, statistic, p.value, protein) %>%
    rename(Log_OR = estimate, SE = std.error, Z = statistic, P = p.value) %>%
    mutate(
      OR = exp(Log_OR),
      CI_Lower = exp(Log_OR - 1.96 * SE),
      CI_Upper = exp(Log_OR + 1.96 * SE),
      Group = group,
      P_adj = p.adjust(P, method = "fdr"),
    )

  all_results[[group]] <- group_clean
  write_csv(group_clean, paste0("~/afib/outputs/proteomics/mono_cancer/logit_", group, ".csv"))
}

# Combine all results
combined_results <- bind_rows(all_results)
write_csv(combined_results, "~/afib/outputs/proteomics/mono_cancer/logit_combined.csv")


# Sample size table for each cancer group
sample_n_df <- working_df %>%
  filter(icd_group %in% cancer_groups | cancer_status == 0) %>%
  mutate(group = if_else(cancer_status == 1, icd_group, "no_cancer")) %>%
  group_by(group) %>%
  summarise(
    n_cancer = sum(cancer_status == 1 & icd_group == group),
    n_control = sum(cancer_status == 0),
    n_total = n_cancer + n_control,
    .groups = "drop"
  ) %>%
  # Add total pan-cancer row
  bind_rows(
    working_df %>%
      filter(cancer_status %in% c(0, 1)) %>%
      summarise(
        group = "pan_cancer",
        n_cancer = sum(cancer_status == 1),
        n_control = sum(cancer_status == 0),
        n_total = n_cancer + n_control
      )
  )

# Plotting
groups <- unique(combined_results$Group)

pdf("~/afib/outputs/proteomics/mono_cancer/mono_cancer_plots.pdf")

for (grp in groups) {
  message("Plotting: ", grp)
  #plot pdf
  
  n_cancer <- sample_n_df$n_cancer[sample_n_df$group == grp]
  n_control <- sample_n_df$n_control[sample_n_df$group=="no_cancer"]

  volcano_df <- combined_results %>%
    filter(Group == grp) %>%
    mutate(
      neg_log_p = -log10(P_adj),
      direction = case_when(
        P_adj >= 0.05 ~ "Neutral",
        Log_OR > 0 ~ "Up (Risk ↑)",
        Log_OR < 0 ~ "Down (Protective)"
      ),
      label = ifelse(P_adj < 0.05 & abs(Log_OR) > 0.25, protein, NA)
    )

  # Plot
  p <- ggplot(volcano_df, aes(x = Log_OR, y = neg_log_p, color = direction)) +
    geom_point(alpha = 0.9, size = 1.5) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "darkred") +
    geom_text_repel(aes(label = label), size = 4, max.overlaps = 25) +
    scale_color_manual(
      values = c(
        "Up (Risk ↑)" = "#D55E00",
        "Down (Protective)" = "#0072B2",
        "Neutral" = "grey70"
      )
    ) +
    labs(
      title = paste("AFib +", grp, "vs", grp, "Only"),
      subtitle = paste0("n = ", n_cancer, " cancer, n = ", n_control, " no cancer"),
      x = "log(Odds Ratio)",
      y = expression(-log[10]~"(FDR-Adjusted P-Value)"),
      color = "Direction"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5)
    ) +
    coord_cartesian(ylim = c(0, 20))

  print(p)
}
dev.off()

```

### Add in columns for proteins found in the mono studies
```{r}

afib_control <- read_csv("~/afib/outputs/proteomics/afib_cancer/afib_control/logit_combined.csv")
cancer_control <- read_csv("~/afib/outputs/proteomics/afib_cancer/cancer_control/logit_combined.csv")

mono_afib <- read_csv("~/afib/outputs/proteomics/mono_afib/cox_results.csv")
mono_cancer <- read_csv("~/afib/outputs/proteomics/mono_cancer/logit_combined.csv")


# Extract list of proteins from mono_afib
mono_afib_proteins <- mono_afib %>%
  filter(P_adj <= 0.05) %>%
  pull(Protein) %>%
  unique()


# Add in_mono_afib column to both datasets
afib_control <- afib_control %>%
  mutate(sig_in_mono_afib = protein %in% mono_afib_proteins)

cancer_control <- cancer_control %>%
  mutate(sig_in_mono_afib = protein %in% mono_afib_proteins)



# do the same for mono_cancer proteins

# Initialize columns
afib_control$sig_in_mono_cancer <- FALSE
cancer_control$sig_in_mono_cancer <- FALSE

# Loop through cancer groups
cancer_groups <- unique(mono_cancer$Group)

for (grp in cancer_groups) {
  # Get significant proteins for this group
  sig_prots <- mono_cancer %>%
    filter(Group == grp, P_adj <= 0.05) %>%
    pull(protein) %>%
    unique()
  
  # Flag in afib_control
  afib_control <- afib_control %>%
    mutate(sig_in_mono_cancer = if_else(
      Group == grp & protein %in% sig_prots,
      TRUE,
      sig_in_mono_cancer
    ))
  
  # Flag in cancer_control
  cancer_control <- cancer_control %>%
    mutate(sig_in_mono_cancer = if_else(
      Group == grp & protein %in% sig_prots,
      TRUE,
      sig_in_mono_cancer
    ))
}


write_csv(afib_control, "~/afib/outputs/proteomics/afib_cancer/afib_control/logit_combined_hits.csv")
write_csv(cancer_control, "~/afib/outputs/proteomics/afib_cancer/cancer_control/logit_combined_hits.csv")


```


### 5-cancer grouping
```{r}
# Define ICD code prefixes for the 5 high-risk cancers
high_risk_prefixes <- c("C15", "C16", "C34", "C90", "C81", "C82", "C83", "C84", "C85", "C86", "C88")

# Create binary column: TRUE if any ICD matches a high-risk prefix
working_df <- working_df %>%
  mutate(
    high_afib_risk_cancer = str_sub(cancer_icd, 1, 3) %in% high_risk_prefixes
  )

# Filter out no cancer
df_sub <- working_df %>%
  filter(cancer_status == 0 | high_afib_risk_cancer == TRUE)


# mono cancer logistic regression

# Covariates
covariates <- c("age", "sex", paste0("PC", 1:10), "ethnicity")
cov_str <- paste(covariates, collapse = " + ")

# Define protein variables (omit eid)
protein_vars <- setdiff(names(olink_df), "eid")


# Logistic regression
highrisk_results <- map_dfr(protein_vars, function(prot) {
  formula_str <- paste0("high_afib_risk_cancer ~ ", prot, " + ", cov_str)
  
  tryCatch({
    fit <- glm(as.formula(formula_str), data = df_sub, family = binomial)
    
    tidy(fit) %>%
      filter(term == prot) %>%
      mutate(protein = prot)
  }, error = function(e) NULL)
})

# Clean + save
highrisk_clean <- highrisk_results %>%
  select(term, estimate, std.error, statistic, p.value, protein) %>%
  rename(Log_OR = estimate, SE = std.error, Z = statistic, P = p.value) %>%
  mutate(
    OR = exp(Log_OR),
    CI_Lower = exp(Log_OR - 1.96 * SE),
    CI_Upper = exp(Log_OR + 1.96 * SE),
    Group = "high_afib_risk_cancer",
    P_adj = p.adjust(P, method = "fdr")
  )

write_csv(
  highrisk_clean,
  "~/afib/outputs/proteomics/mono_cancer/5cancer_group/logit.csv"
)

# Sample sizes
sample_n_df <- working_df %>%
  filter(
    high_afib_risk_cancer == TRUE | cancer_status == 0
  ) %>%
  count(group = high_afib_risk_cancer) %>%
  mutate(
    label = if_else(group, "High-AFib-Risk Cancer", "No Cancer")
  )

# Extract n's
n_cancer <- sample_n_df$n[sample_n_df$group == TRUE]
n_control <- sample_n_df$n[sample_n_df$group == FALSE]

# Prep data
volcano_df <- highrisk_clean %>%
  mutate(
    neg_log_p = -log10(P_adj),
    direction = case_when(
      P_adj >= 0.05 ~ "Neutral",
      Log_OR > 0 ~ "Up (↑ in High-Risk Cancer)",
      Log_OR < 0 ~ "Down (↓ in High-Risk Cancer)"
    ),
    label = ifelse(P_adj < 0.05 & abs(Log_OR) > 0.25, protein, NA)
  )

# Plot
p1 <- ggplot(volcano_df, aes(x = Log_OR, y = neg_log_p, color = direction)) +
  geom_point(alpha = 0.9, size = 1.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "darkred") +
  geom_text_repel(aes(label = label), size = 4, max.overlaps = 25) +
  scale_color_manual(
    values = c(
      "Up (↑ in High-Risk Cancer)" = "#D55E00",
      "Down (↓ in High-Risk Cancer)" = "#0072B2",
      "Neutral" = "grey70"
    )
  ) +
  labs(
    title = "High AFib-Risk Cancer vs No Cancer",
    subtitle = paste0("n = ", n_cancer, " with cancer, n = ", n_control, " without cancer"),
    x = "log(Odds Ratio)",
    y = expression(-log[10]~"(FDR-Adjusted P-Value)"),
    color = "Direction"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_cartesian(ylim = c(0, 20))  # adjust y-limit as needed

print(p1)


# Subset to AFib-only participants
afib_only_df <- df_sub %>% filter(afib_status == 1)

# Logistic regression loop: outcome = high_afib_risk_cancer
afib_control_results <- map_dfr(protein_vars, function(prot) {
  formula_str <- paste0("high_afib_risk_cancer ~ ", prot, " + ", cov_str)
  tryCatch({
    fit <- glm(as.formula(formula_str), data = afib_only_df, family = binomial)
    broom::tidy(fit) %>%
      filter(term == prot) %>%
      mutate(protein = prot)
  }, error = function(e) NULL)
})

# Clean up and annotate
afib_control_clean <- afib_control_results %>%
  select(term, estimate, std.error, statistic, p.value, protein) %>%
  rename(Log_OR = estimate, SE = std.error, Z = statistic, P = p.value) %>%
  mutate(
    OR = exp(Log_OR),
    CI_Lower = exp(Log_OR - 1.96 * SE),
    CI_Upper = exp(Log_OR + 1.96 * SE),
    Group = "highrisk_cancer",
    P_adj = p.adjust(P, method = "fdr")
  )

# Save the table
readr::write_csv(afib_control_clean, "~/afib/outputs/proteomics/afib_cancer/afib_control/5cancer_group/logit.csv")


# Sample sizes for subtitle
sample_n_df <- afib_only_df %>%
  filter(!is.na(high_afib_risk_cancer)) %>%
  count(group = high_afib_risk_cancer) %>%
  mutate(
    label = if_else(group, "High-AFib-Risk Cancer", "AFib Only")
  )

n_cancer <- sample_n_df$n[sample_n_df$group == TRUE]
n_afib_only <- sample_n_df$n[sample_n_df$group == FALSE]

volcano_df <- afib_control_clean %>%
  mutate(
    neg_log_p = -log10(P_adj),
    direction = case_when(
      P_adj >= 0.05 ~ "Neutral",
      Log_OR > 0 ~ "Up (↑ in High-Risk Cancer)",
      Log_OR < 0 ~ "Down (↓ in High-Risk Cancer)"
    ),
    label = ifelse(P_adj < 0.05 & abs(Log_OR) > 0.25, protein, NA)
  )



p2 <- ggplot(volcano_df, aes(x = Log_OR, y = neg_log_p, color = direction)) +
  geom_point(alpha = 0.9, size = 1.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "darkred") +
  geom_text_repel(aes(label = label), size = 4, max.overlaps = 25) +
  scale_color_manual(
    values = c(
      "Up (↑ in High-Risk Cancer)" = "#D55E00",
      "Down (↓ in High-Risk Cancer)" = "#0072B2",
      "Neutral" = "grey70"
    )
  ) +
  labs(
    title = "AFib + High-Risk Cancer vs AFib Only",
    subtitle = paste0("n = ", n_cancer, " with high-risk cancer, n = ", n_afib_only, " AFib-only"),
    x = "log(Odds Ratio)",
    y = expression(-log[10]~"(FDR-Adjusted P-Value)"),
    color = "Direction"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_cartesian(ylim = c(0, 20))  # Adjust as needed

print(p2)



# Subset to those with cancer
cancer_only_df <- working_df %>%
  filter(high_afib_risk_cancer == TRUE)


# Run logistic regression: outcome = afib_status
cancer_control_results <- map_dfr(protein_vars, function(prot) {
  formula_str <- paste0("afib_status ~ ", prot, " + ", cov_str)
  tryCatch({
    fit <- glm(as.formula(formula_str), data = cancer_only_df, family = binomial)
    broom::tidy(fit) %>%
      filter(term == prot) %>%
      mutate(protein = prot)
  }, error = function(e) NULL)
})

# Clean and format
cancer_control_clean <- cancer_control_results %>%
  select(term, estimate, std.error, statistic, p.value, protein) %>%
  rename(Log_OR = estimate, SE = std.error, Z = statistic, P = p.value) %>%
  mutate(
    OR = exp(Log_OR),
    CI_Lower = exp(Log_OR - 1.96 * SE),
    CI_Upper = exp(Log_OR + 1.96 * SE),
    Group = "highrisk_cancer",
    P_adj = p.adjust(P, method = "fdr")
  )

# Save
readr::write_csv(cancer_control_clean, "~/afib/outputs/proteomics/afib_cancer/cancer_control/5cancer_group/logit.csv")



# n's for cancer population: AFib vs No AFib
sample_n_df <- cancer_only_df %>%
  count(group = afib_status) %>%
  mutate(
    label = if_else(group == 1, "AFib+", "Cancer-only")
  )

n_afib <- sample_n_df$n[sample_n_df$group == 1]
n_cancer_only <- sample_n_df$n[sample_n_df$group == 0]


# Prepare volcano dataframe
volcano_df <- cancer_control_clean %>%
  mutate(
    neg_log_p = -log10(P_adj),
    direction = case_when(
      P_adj >= 0.05 ~ "Neutral",
      Log_OR > 0 ~ "Up (↑ in AFib + Cancer)",
      Log_OR < 0 ~ "Down (↓ in AFib + Cancer)"
    ),
    label = ifelse(P_adj < 0.05 & abs(Log_OR) > 0.25, protein, NA)
  )


p3 <- ggplot(volcano_df, aes(x = Log_OR, y = neg_log_p, color = direction)) +
  geom_point(alpha = 0.9, size = 1.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "darkred") +
  geom_text_repel(aes(label = label), size = 4, max.overlaps = 25) +
  scale_color_manual(
    values = c(
      "Up (↑ in AFib + Cancer)" = "#D55E00",
      "Down (↓ in AFib + Cancer)" = "#0072B2",
      "Neutral" = "grey70"
    )
  ) +
  labs(
    title = "AFib + Cancer vs Cancer Only",
    subtitle = paste0("n = ", n_afib, " with AFib, n = ", n_cancer_only, " cancer-only"),
    x = "log(Odds Ratio)",
    y = expression(-log[10]~"(FDR-Adjusted P-Value)"),
    color = "Direction"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_cartesian(ylim = c(0, 20))  # adjust if needed

print(p3)



# Add in the columns
mono_afib <- read_csv("~/afib/outputs/proteomics/mono_afib/cox_results.csv")

afib_control <- afib_control_clean
cancer_control <- cancer_control_clean

# Filter significant mono-AFib proteins
mono_afib_proteins <- mono_afib %>%
  filter(P_adj <= 0.05) %>%
  pull(Protein) %>%
  unique()

# Add sig_in_mono_afib flag
afib_control <- afib_control %>%
  mutate(sig_in_mono_afib = protein %in% mono_afib_proteins)

cancer_control <- cancer_control %>%
  mutate(sig_in_mono_afib = protein %in% mono_afib_proteins)



# Get sig mono-cancer proteins by group
sig_mono_cancer_prots <- highrisk_clean %>%
  filter(P_adj <= 0.05) %>%
  select(Group, protein) %>%
  mutate(sig_in_mono_cancer = TRUE)

# Add to afib_control
afib_control <- afib_control %>%
  left_join(sig_mono_cancer_prots, by = c("Group", "protein")) %>%
  mutate(sig_in_mono_cancer = replace_na(sig_in_mono_cancer, FALSE))

# Add to cancer_control
cancer_control <- cancer_control %>%
  left_join(sig_mono_cancer_prots, by = c("Group", "protein")) %>%
  mutate(sig_in_mono_cancer = replace_na(sig_in_mono_cancer, FALSE))

readr::write_csv(afib_control, "~/afib/outputs/proteomics/afib_cancer/afib_control/5cancer_group/logit.csv")
readr::write_csv(cancer_control, "~/afib/outputs/proteomics/afib_cancer/cancer_control/5cancer_group/logit.csv")



```

### Forest plots for manuscript figure
```{r}
# Load results
lung <- read_csv("~/afib/outputs/proteomics/afib_cancer/afib_control/logit_lung.csv") %>%
  mutate(Stratum = "AF+Lung")
nhl <- read_csv("~/afib/outputs/proteomics/afib_cancer/afib_control/logit_nhl.csv") %>%
  mutate(Stratum = "AF+NHL")
pancancer <- read_csv("~/afib/outputs/proteomics/afib_cancer/afib_control/logit_pancancer.csv") %>%
  mutate(Stratum = "AF+Pan-Cancer")

# Combine and filter to four proteins
target_proteins <- c("cd200r1", "duox2", "itgam", "clec5a")

# Build forest_df with only relevant strata per protein
forest_df <- bind_rows(
  lung %>% filter(protein %in% c("cd200r1", "duox2")) %>%
    mutate(Stratum = "AF+Lung"),
  nhl %>% filter(protein %in% c("itgam", "clec5a")) %>%
    mutate(Stratum = "AF+NHL"),
  pancancer %>% filter(protein %in% c("cd200r1", "duox2", "itgam", "clec5a")) %>%
    mutate(Stratum = "AF+Pan-Cancer")
) %>%
  select(protein, Stratum, OR, CI_Lower, CI_Upper, P_adj) %>%
  mutate(
    protein = factor(protein,
      levels = c("cd200r1", "duox2", "itgam", "clec5a"),
      labels = c("CD200R1", "DUOX2", "ITGAM", "CLEC5A")
    ),
    Stratum = factor(Stratum, levels = c("AF+Lung", "AF+NHL", "AF+Pan-Cancer")),
    Direction = case_when(
      CI_Lower > 1 ~ "Positive",
      CI_Upper < 1 ~ "Inverse",
      TRUE ~ "Non-significant"
    )
  )

# Plot
ggplot(forest_df, aes(x = OR, y = Stratum, color = Direction)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(
    aes(xmin = CI_Lower, xmax = CI_Upper),
    size = 0.8,
    fatten = 3
  ) +
  scale_color_manual(values = c(
    "Positive" = "#D55E00",
    "Inverse" = "#0072B2",
    "Non-significant" = "grey60"
  )) +
  facet_wrap(~ protein, ncol = 2, scales = "free_y") +
  labs(
    x = "Odds Ratio (95% CI)",
    y = NULL,
    color = "Direction of Association"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    axis.text.y = element_text(size = 11)
  )


# Save as TIFF (most universally accepted)
ggsave(
  "~/afib/outputs/proteomics/afib_cancer/afib_control/forest_plot.tiff",
  width = 7,
  height = 5,
  units = "in",
  dpi = 600,
  compression = "lzw"
)

ggsave(
  "~/afib/outputs/proteomics/afib_cancer/afib_control/forest_plot.pdf",
  width = 7,
  height = 5,
  units = "in"
)

ggsave(
  "~/afib/outputs/proteomics/afib_cancer/afib_control/forest_plot.png",
  width = 7,
  height = 5,
  units = "in",
  dpi = 600
)
```


